Precise spiking motifs in neurobiological and neuromorphic data

### 2022-12-16 manuscript.tex
Update on Overleaf.

\add[Antoine]{In this first section, we have introduced the notion of rate coding and demonstrated that spike timing can also carry information. Scientists found experimental evidence for various hypotheses of neural representations like population coding}~\cite{Maunsell_Functional_1983}\add[Antoine]{, time-to-first-spike coding}~\citep{gollisch_rapid_2008}\add[Antoine]{, phase-of-firing coding}~\citep{montemurro_phase-of-firing_2008}\add[Antoine]{, correlation coding}~\citep{decharms_primary_1996}\add[Antoine]{ or sparse coding}~\citep{vinje_sparse_2000}\add[Antoine]{. In the scope of this review, we infer that spike trains are composed of repeating spiking motifs and we focus on precise spatio-temporal representations composed of a motif of spikes defined precisely in time and in the presynaptic address space. In all generality, this representation can encompass all the previous ones if the precision in time . }

@@

In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that neural assemblies could form so-called ``synfire chains,'' that is, showing the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018} and were also extended to larger assemblies using statistical methods~\citep{torre_asset_2016} (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intracortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active, allowing it to more efficiently drive cortical neurons~\citep{bruno_cortex_2006}. \noet{neuron-centric view from the reviewer comment, make some changes to broaden the proposed vision. Also change the very specific example given after, or put in context (works by Sophie Dénève or Ballard and Jehee)}

@@

Propagating waves in the neuronal response occur in many excitable media and were recently found in neural systems, for instance in the retina or the neocortex. While propagating waves are clearly present under anesthesia, whether they also appear during awake and conscious states remained unclear until recent discoveries. One possibility is that these waves were systematically missed in trial-averaged data, due to variability. A recent work~\citep{muller_stimulus-evoked_2014} presents a method for detecting propagating waves in noisy multichannel recordings. Applying this method to single-trial voltage-sensitive dye imaging data, the authors show that the stimulus-evoked population response in the primary visual cortex of the awake monkey propagates as a travelling wave, with consistent dynamics across trials. A network model suggests that this reliability is the hallmark of the horizontal fiber network of superficial cortical layers. Propagating waves with similar properties occur independently in the secondary visual cortex, but maintain precise phase relations with the waves in the primary visual cortex. These results show that, in response to a visual stimulus, propagating waves are systematically evoked in several visual areas, generating a consistent spatio-temporal frame for further neural interactions.

@@

To our knowledge, the first neuromorphic circuit is the pulsed current-source synapse proposed by Carver Mead in 1989~\citep{mead_analog_1989}. It was implemented with transistors operating in the sub-threshold domain and responded to asynchronous events, but was not capable of discriminating two different spiking sequences with the same firing rate. Indeed, the postsynaptic membrane potential was increased by a step proportional to the input current but did not decrease in time, as it can be observed in biological neurons. Then, electronics circuits became more and more bio-realistic and, two decades later, ~\citep{bartolozzi_synaptic_2007} released the Diff-Pair Integrator (DPI) synapse that could reproduce the global dynamics of the biological neurons. The DPI circuit could multiplex in time spikes from different sources and became a potential  ``silicon coincidence detector''. Today, many devices are good candidates for implementing event-based algorithms and use the \change[Antoine]{AER}{address event representation (AER)}. They can be divided intro 3 major categories: digital, analog and mixed analog/digital platforms. For a more complete review the reader can refer to~\citep{schuman_survey_2017}, here we site the most popular ones. SpiNNaker~\citep{furber_overview_2013, furber_spinnaker_2020}, Loihi~\citep{davies_loihi_2018} and TrueNorth~\citep{merolla_million_2014} chips are widely used examples of digital hardware implementations. Compared to TrueNorth which exclusively implements a LIF neuron, SpiNNaker and Loihi offer some flexibility in terms of neuron model and allow for on-chip learning. This flexibility in the implementation comes at the cost of an increased energy consumption. Mixed-analog-digital systems were developed at the Stanford University: Neurogrid and Braindrop~\citep{benjamin_neurogrid_2014, neckar_braindrop_2019}. They are mostly used by computational neuroscientist to model brain activity with different levels of abstraction. BrainScaleS~\citep{schemmel_wafer-scale_2010} is another mixed-analog-digital system developed, just like SpiNNaker, for the Human Brain Project~\citep{markram_introducing_2011}. It is a wafer-scale neuromorphic hardware with analog components. Analog arrays (i.e. field programmable analog arrays (FPAA)) refer to the initial idea of neuromorphic hardware aiming at building strictly analog devices. The pulsed current source synapse and the DPI are examples of such devices, we can also mention the field programmable neural array~\citep{farquhar_field_2006} and the NeuroFPAA~\citep{cheng_fpaa_2009} specifically designed for neuromorphic systems. Due to their lack of generality and some issues specific to analog circuits, these fully analog devices are not yet widely used for neuromorphic computing. 

>>>>

\add[Antoine]{In this first section, we have introduced the notion of rate coding and demonstrated that spike timing can also carry information. Scientists found experimental evidence for various hypotheses of neural representations like population coding}~\cite{Maunsell_Functional_1983}\add[Antoine]{, time-to-first-spike coding}~\citep{gollisch_rapid_2008}\add[Antoine]{, phase-of-firing coding}~\citep{montemurro_phase-of-firing_2008}\add[Antoine]{, correlation coding}~\citep{decharms_primary_1996}\add[Antoine]{ or sparse coding}~\citep{vinje_sparse_2000}\add[Antoine]{. In the scope of this review, we infer that spike trains are composed of repeating spiking motifs and we focus on precise spatio-temporal representations composed of a motif of spikes defined precisely in time and in the presynaptic address space. In all generality, this representation can encompass all the previous ones except for rate coding which is not defined locally in time. In this section, we choose to described different hypotheses making use of spatio-temporal patterns of spikes that can be propagated among neural assemblies.}

@@

In his book ``Corticonics'',~\citet{abeles_corticonics_1991} queried whether the role of cortical neurons is to integrate synaptic inputs or rather to detect coincidences in temporal spiking patterns. The book gradually leads the reader from the macroscopic cortical anatomy and standard electro-physiological properties of single neurons to neural network models. While the first hypothesis favors the rate coding theory, the second possibility highlights the need for temporal precision in the neural code~\citep{abeles_role_1982,paugam-moisy_computing_2012}. The book then demonstrates that neural assemblies could form so-called ``synfire chains,'' that is, showing the emergence of synchronous activity on subsets of neurons which could be propagated in a stable fashion. More broadly, the idea of using the synchrony of co-activation in a cell assembly is reminiscent of the hypothesis that was formalized by~\citet{hebb_organization_1949}: \add[Antoine]{"cells that fire together wire together"}. Since this date, multiple experimental observations have suggested the existence of this precise zero-phase-lag spike synchronization in a defined subset of neurons~\citep{harris_organization_2003}. One possible function of this synchronization may serve the binding of information distributed in the brain~\citep{singer_visual_1995, roelfsema_visuomotor_1997}.

Some experimental results show the emergence of synchrony, for instance in motor cortical function~\citep{riehle_spike_1997}. Interestingly, these authors showed that ``accurate spike synchronization occurred in relation to external events (stimuli, movements) and was commonly accompanied by discharge rate modulations but without precise time locking of the spikes to these external events. Spike synchronization also occurred in relation to purely internal events (stimulus expectancy), where firing rate modulations were distinctly absent. These findings indicate that internally generated synchronization of individual spike discharges may subserve the cortical organization of cognitive motor processes.'' Moreover, such emergence could change over the learning period involved in learning a task~\citep{kilavik_long-term_2009} and showed some tuning to movement direction and reaction time~\citep{grammont_spike_2003}. It is important to note that synchronous events tend to lock to spatio-temporal patterns of neural activity called \href{https://en.wikipedia.org/wiki/Local_field_potential}{LFP} beta waves~\citep{denker_lfp_2018} and were also extended to larger assemblies using statistical methods~\citep{torre_asset_2016} (see Section~\ref{sec:detection} for further details). Note also that synchronicity may explain some unintuitive results. Indeed, it has been shown that thalamo-cortical synapses are relatively weak compared to the amount of intracortical activity. However, this pathway is sufficient to drive the cortex, as this input is more often synchronously active, allowing it to more efficiently drive cortical neurons~\citep{bruno_cortex_2006}.

@@

Propagating waves in the neuronal response occur in many excitable media and were recently found in neural systems, for instance in the retina~\citep{feller_dynamic_1997} or the neocortex. While propagating waves are clearly present under anesthesia, whether they also appear during awake and conscious states remained unclear until recent discoveries. One possibility is that these waves were systematically missed in trial-averaged data, due to variability. A recent work~\citep{muller_stimulus-evoked_2014} presents a method for detecting propagating waves in noisy multichannel recordings. Applying this method to single-trial voltage-sensitive dye imaging data, the authors show that the stimulus-evoked population response in the primary visual cortex of the awake monkey propagates as a travelling wave, with consistent dynamics across trials. A network model suggests that this reliability is the hallmark of the horizontal fiber network of superficial cortical layers. Propagating waves with similar properties occur independently in the secondary visual cortex, but maintain precise phase relations with the waves in the primary visual cortex. These results show that, in response to a visual stimulus, propagating waves are systematically evoked in several visual areas, generating a consistent spatio-temporal frame for further neural interactions.

@@

To our knowledge, the first neuromorphic circuit is the pulsed current-source synapse proposed by Carver Mead in 1989~\citep{mead_analog_1989}. It was implemented with transistors operating in the sub-threshold domain and responded to asynchronous events, but was not capable of discriminating two different spiking sequences with the same firing rate. Indeed, the postsynaptic membrane potential was increased by a step proportional to the input current but did not decrease in time, as it can be observed in biological neurons. Then, electronics circuits became more and more bio-realistic and, two decades later, ~\citep{bartolozzi_synaptic_2007} released the Diff-Pair Integrator (DPI) synapse that could reproduce the global dynamics of the biological neurons. The DPI circuit could multiplex in time spikes from different sources and became a potential  ``silicon coincidence detector''. Today, many devices are good candidates for implementing event-based algorithms and use the \change[Antoine]{AER}{address event representation (AER)}. They can be divided intro 3 major categories: digital, analog and mixed analog/digital platforms. For a more complete review the reader can refer to~\citep{schuman_survey_2017}, here we site the most popular ones. SpiNNaker~\citep{furber_overview_2013, furber_spinnaker_2020}, Loihi~\citep{davies_loihi_2018} and TrueNorth~\citep{merolla_million_2014} chips are widely used examples of digital hardware implementations. Compared to TrueNorth which exclusively implements a LIF neuron, SpiNNaker and Loihi offer some flexibility in terms of neuron model and allow for on-chip learning. This flexibility in the implementation comes at the cost of an increased energy consumption. Mixed-analog-digital systems were developed at the Stanford University: Neurogrid and Braindrop~\citep{benjamin_neurogrid_2014, neckar_braindrop_2019}. They are mostly used by computational neuroscientists to model brain activity with different levels of abstraction. BrainScaleS~\citep{schemmel_wafer-scale_2010} is another mixed-analog-digital system developed, just like SpiNNaker, for the Human Brain Project~\citep{markram_introducing_2011}. It is a wafer-scale neuromorphic hardware with analog components. Analog arrays (i.e. field programmable analog arrays (FPAA)) refer to the initial idea of neuromorphic hardware aiming at building strictly analog devices. The pulsed current source synapse and the DPI are examples of such devices, we can also mention the field programmable neural array~\citep{farquhar_field_2006} and the NeuroFPAA~\citep{cheng_fpaa_2009} specifically designed for neuromorphic systems. Due to their lack of generality and some issues specific to analog circuits, these fully analog devices are not yet widely used for neuromorphic computing. 
